---
title: "Cross-Validation for Model Selection"
subtitle: "Evaluating Polynomial Regression Models on the Auto Dataset"
author: "Saroj Bhatta"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    highlight-style: github
    code-fold: false
    code-tools: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Cross-validation is a resampling technique used to estimate how well a statistical model generalizes to an independent dataset. Rather than evaluating a model on the same data used to train it — which produces overly optimistic performance estimates — cross-validation holds out portions of the data to simulate out-of-sample performance.

This report applies the **validation set approach** and **5-fold cross-validation** to select between three polynomial regression models for predicting fuel efficiency (`mpg`) from engine horsepower using the `Auto` dataset.

---

## Background: K-Fold Cross-Validation

In k-fold CV, the dataset is randomly partitioned into *K* equal-sized folds. The model is trained on K−1 folds and evaluated on the held-out fold, rotating through all K folds. The final performance estimate is the average RMSE across all K iterations.

**Comparison with alternatives:**

- *Validation Set Approach:* Simple, but produces high-variance estimates — results depend heavily on the random split.
- *LOOCV (Leave-One-Out CV):* Nearly unbiased, but computationally expensive and can yield high variance across iterations.
- *K-fold CV:* Strikes a practical balance between bias, variance, and computation time.

---

## Setup

```{r}
#| label: load-packages

library(tidyverse)
library(tidymodels)
library(ISLR)
```

---

## Dataset Overview

The `Auto` dataset from the ISLR package contains fuel efficiency and technical specifications for 392 automobiles. The outcome of interest is `mpg` (miles per gallon); the primary predictor is `horsepower`.

```{r}
#| label: explore-data

glimpse(Auto)
```

```{r}
#| label: check-missing

# Confirm no missing values
sum(is.na(Auto))

# drop_na() as further confirmation
auto_clean <- Auto |> drop_na()
glimpse(auto_clean)
```

The dataset has **392 rows** and **9 columns** with **no missing values**, confirmed both by `sum(is.na(Auto))` returning 0 and by `drop_na()` leaving row count unchanged.

---

## Exploratory Analysis

### Distribution of MPG

```{r}
#| label: fig-mpg-hist
#| fig-cap: "Distribution of miles per gallon across 392 vehicles."

ggplot(Auto, aes(x = mpg)) +
  geom_histogram(bins = 30, color = 'skyblue') +
  labs(
    title = 'Distribution of Miles Per Gallon (MPG)',
    x = 'MPG',
    y = 'Count',
    caption = "Source: Auto dataset from ISLR package"
  )
```

```{r}
#| label: fig-mpg-density
#| fig-cap: "MPG distribution with overlaid kernel density estimate."

ggplot(Auto, aes(x = mpg)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = 'skyblue', color = 'black') +
  geom_density(alpha = 0.5, fill = 'darkblue') +
  labs(
    title = 'Distribution of Miles Per Gallon (MPG)',
    x = 'MPG',
    y = 'Density',
    caption = "Source: Auto dataset from ISLR package"
  )
```

The distribution of MPG is right-skewed, with most vehicles between 15–30 MPG and a smaller tail of more fuel-efficient vehicles.

### MPG vs. Horsepower

```{r}
#| label: fig-hp-mpg
#| fig-cap: "Relationship between horsepower and MPG with linear regression smoother."

ggplot(Auto, aes(x = horsepower, y = mpg)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", formula = y ~ x, color = "darkred") +
  labs(
    title = "Relationship between Horsepower and MPG",
    x = "Horsepower",
    y = "MPG"
  )

# A loess smoother can also be added to visualize non-linearity:
# geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
#             color = "red", linetype = "dashed")
```

The scatterplot shows a clear **negative, non-linear** relationship: higher-horsepower vehicles achieve lower fuel efficiency. The curvature suggests a polynomial model may outperform a simple linear fit.

---

## Validation Set Approach

The data was split 75/25 into training (n = 294) and testing (n = 98) sets with a fixed seed for reproducibility. Three polynomial regression models were fit on the training data and evaluated on the test set.

```{r}
#| label: train-test-split

set.seed(42)
auto_split <- initial_split(data = Auto, prop = 3/4)
Auto_train <- training(auto_split)
Auto_test  <- testing(auto_split)

auto_split
```

```{r}
#| label: fit-models-validation

lm_spec <- linear_reg() |> set_engine("lm")

# Model 1: Linear regression
lm1_results <- last_fit(lm_spec,
                        mpg ~ horsepower,
                        split = auto_split)
lm1_results |> collect_metrics()
```

```{r}
# Model 2: Quadratic regression
lm2_results <- last_fit(lm_spec,
                        mpg ~ horsepower + I(horsepower^2),
                        split = auto_split)
lm2_results |> collect_metrics()
```

```{r}
# Model 3: Cubic regression
lm3_results <- last_fit(lm_spec,
                        mpg ~ horsepower + I(horsepower^2) + I(horsepower^3),
                        split = auto_split)
lm3_results |> collect_metrics()
```

### Alternative: Manual Fit and Predict

These results can also be obtained by fitting and predicting separately, without `last_fit()`:

```{r}
#| label: manual-rmse

lm1_fit <- fit(lm_spec, mpg ~ horsepower, data = Auto_train)
pred1 <- lm1_fit |> predict(new_data = Auto_test) |> bind_cols(Auto_test)
rmse1 <- pred1 |> rmse(truth = mpg, estimate = .pred)
rmse1

lm2_fit <- fit(lm_spec, mpg ~ poly(horsepower, 2), data = Auto_train)
pred2 <- lm2_fit |> predict(new_data = Auto_test) |> bind_cols(Auto_test)
rmse2 <- pred2 |> rmse(truth = mpg, estimate = .pred)
rmse2

lm3_fit <- fit(lm_spec, mpg ~ poly(horsepower, 3), data = Auto_train)
pred3 <- lm3_fit |> predict(new_data = Auto_test) |> bind_cols(Auto_test)
rmse3 <- pred3 |> rmse(truth = mpg, estimate = .pred)
rmse3
```

### Validation Set Results Summary

| Model | Formula | Test RMSE | Test R² |
|-------|---------|-----------|---------|
| Model 1 | mpg ~ horsepower | 5.11 | 0.617 |
| Model 2 | mpg ~ horsepower + hp² | 4.56 | 0.700 |
| Model 3 | mpg ~ horsepower + hp² + hp³ | 4.56 | 0.701 |

Model 2 and Model 3 achieve the same test RMSE (4.56). Adding the cubic term provides no meaningful improvement. Applying the principle of **parsimony**, **Model 2 (quadratic)** is preferred.

---

## 5-Fold Cross-Validation

To obtain more reliable performance estimates, the same three models were re-evaluated using 5-fold CV on the full dataset.

```{r}
#| label: kfold-cv

set.seed(143)
auto_cv <- vfold_cv(Auto, 5)
auto_cv
```

```{r}
#| label: cv-results

# Model 1
results1 <- fit_resamples(lm_spec, mpg ~ horsepower, resamples = auto_cv)
results1 |> collect_metrics()
```

```{r}
# Model 2
results2 <- fit_resamples(lm_spec, mpg ~ poly(horsepower, 2), resamples = auto_cv)
results2 |> collect_metrics()
```

```{r}
# Model 3
results3 <- fit_resamples(lm_spec, mpg ~ poly(horsepower, 3), resamples = auto_cv)
results3 |> collect_metrics()
```

| Model | Mean CV RMSE | Mean CV R² |
|-------|-------------|------------|
| Model 1 (Linear) | 4.95 | 0.618 |
| Model 2 (Quadratic) | 4.38 | 0.687 |
| Model 3 (Cubic) | 4.39 | 0.686 |

5-fold CV confirms the validation set findings: **Model 2 achieves the lowest CV RMSE (4.38)**. The cubic term again adds no improvement.

---

## Automated Tuning with `tune_grid()`

Rather than manually fitting models at fixed degrees, `step_poly(horsepower, degree = tune())` within a tidymodels recipe automates the search. This step creates orthogonal polynomial basis expansions of the predictor, treating the degree as a hyperparameter to be optimized via cross-validation.

```{r}
#| label: auto-tune

auto_prep <- Auto |>
  recipe(mpg ~ horsepower) |>
  step_poly(horsepower, degree = tune())

auto_tune <- tune_grid(lm_spec, auto_prep, resamples = auto_cv)
```

```{r}
#| label: tune-metrics

auto_metrics <- auto_tune |> collect_metrics()
auto_metrics
```

```{r}
#| label: fig-rmse-degree
#| fig-cap: "Mean 5-fold CV RMSE across polynomial degrees with ±1 standard error bars."

auto_rmse <- auto_metrics |> filter(.metric == "rmse")
auto_rmse

ggplot(auto_rmse, aes(x = factor(degree), y = mean)) +
  geom_line(group = 1) +
  geom_pointrange(aes(ymin = mean - std_err, ymax = mean + std_err)) +
  labs(
    x = "Degree",
    y = "Cross-validation error (RMSE)",
    title = "Mean CV RMSE Error vs Polynomial Degree"
  )
```

The plot confirms that **degree 2 minimizes CV error**. Degree 1 performs substantially worse; degree 3 offers only marginal and statistically insignificant improvement over degree 2. The overlapping error bars between degrees 2 and 3 reinforce that the cubic term is not justified.

---

## Conclusions

Across both the validation set approach and 5-fold cross-validation, a **quadratic (degree-2) polynomial** of horsepower provides the best model for predicting MPG:

- Test RMSE drops from **5.11** (linear) to **4.56** (quadratic)
- CV RMSE drops from **4.95** (linear) to **4.38** (quadratic)
- The cubic model adds no practical value — nearly identical RMSE with an extra parameter

This analysis demonstrates the core benefit of cross-validation: it guards against both underfitting and overfitting by providing reliable out-of-sample performance estimates without relying on a single arbitrary data split.
