---
title: "Ridge, Lasso & Elastic Net Regression"
subtitle: "Regularized Regression for Geographic Music Origin Prediction"
author: "Saroj Bhatta"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    highlight-style: github
    code-fold: false
    code-tools: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

When the number of predictors is large relative to the number of observations, ordinary least squares (OLS) tends to overfit. **Regularized regression** addresses this by adding a penalty term to the OLS objective that shrinks coefficient magnitudes — trading a small increase in bias for a potentially large reduction in variance.

This report compares three regularization methods — **ridge regression**, **lasso**, and **elastic net** — applied to predicting the geographic latitude of a music track's origin from 72 audio feature variables. Hyperparameters are selected via 10-fold cross-validation.

---

## Background

| Method | Penalty | Effect |
|--------|---------|--------|
| **Ridge** | L2 (sum of squared coefficients) | Shrinks all coefficients toward zero; none set exactly to zero |
| **Lasso** | L1 (sum of absolute coefficients) | Shrinks coefficients and performs variable selection (some set to exactly zero) |
| **Elastic Net** | L1 + L2 combined | Interpolates between ridge and lasso via the `mixture` parameter (α) |

The penalty strength λ controls how much shrinkage is applied. When λ = 0, all three methods reduce to OLS.

---

## Setup

```{r}
#| label: load-packages

library(tidyverse)
library(tidymodels)
```

---

## Data & Train/Test Split

The music dataset contains 500 observations and 73 variables — 72 audio features plus the outcome `lat` (latitude of music origin). The data was split 75/25 with seed 7 for reproducibility.

```{r}
#| label: load-split-data

set.seed(7)

music <- read_csv("data/music.csv", show_col_types = FALSE)

music_split <- initial_split(music, prop = 3/4)
music_train <- training(music_split)
music_test  <- testing(music_split)

dim(music_train)
dim(music_test)
```

The training set has **375 observations** and 73 variables; the test set has **125 observations**.

---

## Exploratory Analysis: Outcome Distribution

```{r}
#| label: fig-lat-dist
#| fig-cap: "Distribution of latitude (outcome variable) in the training set, showing the geographic spread of music origins."

ggplot(data = music_train, aes(x = lat)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = 'skyblue', color = 'blue') +
  geom_density(fill = "lightgreen", alpha = 0.6) +
  labs(
    title = 'Distribution of Latitude of Music Origin',
    x = 'Latitude',
    y = 'Count / Density'
  )
```

---

## Baseline: Ordinary Least Squares

An OLS model using all 72 predictors serves as the benchmark.

```{r}
#| label: ols-baseline

lm_spec <- linear_reg() |> set_engine("lm")

lm_fit <- fit(lm_spec, lat ~ ., data = music_train)

preds <- lm_fit |>
  predict(new_data = music_test) |>
  bind_cols(music_test)

rmse_ols <- preds |> rmse(truth = lat, estimate = .pred)
rmse_ols
```

The OLS test RMSE is **17.03** — this is the benchmark to beat.

---

## Ridge Regression

Ridge regression adds an L2 penalty: all predictors are retained but their coefficients are shrunk toward zero.

### Preprocessing Recipe

Predictors are dummy-encoded and scaled before fitting. Scaling is essential for ridge regression because the penalty is applied uniformly across coefficients — unscaled predictors with large magnitudes would otherwise receive disproportionate shrinkage.

```{r}
#| label: ridge-recipe

music_cv <- vfold_cv(data = music_train, v = 10)

reg_recipe <- recipe(lat ~ ., data = music_train) |>
  step_dummy(all_nominal()) |>
  step_scale(all_predictors())
```

### Tuning λ via 10-Fold CV

```{r}
#| label: ridge-tune

ridge_spec <- linear_reg(penalty = tune(), mixture = 0) |>
  set_engine("glmnet")

ridge_grid <- expand_grid(penalty = seq(0, 100, by = 10))

ridge_results <- tune_grid(ridge_spec,
                           preprocessor = reg_recipe,
                           grid = ridge_grid,
                           resamples = music_cv)

ridge_results |> collect_metrics()
```

```{r}
#| label: ridge-best-lambda

# Select lambda with lowest CV RMSE
best_rmse <- ridge_results |> select_best(metric = "rmse")
best_lambda <- best_rmse$penalty
best_rmse
```

```{r}
#| label: ridge-rmse-sorted

ridge_results |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  arrange(mean)
```

### Final Ridge Model on Test Set

```{r}
#| label: ridge-final-fit

final_reg_spec <- linear_reg(penalty = best_lambda, mixture = 0) |>
  set_engine("glmnet")

ridge_wf <- workflow() |>
  add_model(final_reg_spec) |>
  add_recipe(reg_recipe)

ridge_fit <- last_fit(ridge_wf, split = music_split)
ridge_fit |> collect_metrics()
```

The optimal λ selected by 10-fold CV is **λ = 10**, yielding a CV RMSE of ~17.06 and a test RMSE of **16.54** — an improvement over the OLS baseline of 17.03.

### Variable Coefficients in Ridge

In ridge regression, all 72 predictors are retained. Coefficients are shrunk toward zero but none are set exactly to zero.

```{r}
#| label: ridge-coefficients

# Fit final ridge model to full training data
final_fit <- workflow() |>
  add_recipe(reg_recipe) |>
  add_model(final_reg_spec) |>
  fit(data = music_train)

final_fit |>
  tidy() |>
  filter(term != "(Intercept)") |>
  arrange(desc(abs(estimate)))
```

---

## Lasso Regression

Lasso uses an L1 penalty, which has the geometric property of setting some coefficients to exactly zero — performing automatic **variable selection**. The specification is identical to ridge but with `mixture = 1`.

```{r}
#| label: lasso-tune

lasso_spec <- linear_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")

lasso_results <- tune_grid(lasso_spec,
                           preprocessor = reg_recipe,
                           grid = ridge_grid,
                           resamples = music_cv)

best_lasso <- lasso_results |> select_best(metric = "rmse")
best_lasso
```

```{r}
#| label: lasso-final

final_lasso_spec <- linear_reg(penalty = best_lasso$penalty, mixture = 1) |>
  set_engine("glmnet")

lasso_wf <- workflow() |>
  add_model(final_lasso_spec) |>
  add_recipe(reg_recipe)

lasso_fit <- last_fit(lasso_wf, split = music_split)
lasso_fit |> collect_metrics()
```

---

## Elastic Net

Elastic net combines L1 and L2 penalties. The `mixture` parameter α interpolates between pure ridge (α = 0) and pure lasso (α = 1). Both `penalty` and `mixture` are tuned jointly.

```{r}
#| label: elastic-net-tune

enet_spec <- linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet")

enet_grid <- expand_grid(
  penalty  = seq(0, 100, by = 10),
  mixture  = c(0, 0.25, 0.5, 0.75, 1)
)

enet_results <- tune_grid(enet_spec,
                          preprocessor = reg_recipe,
                          grid = enet_grid,
                          resamples = music_cv)

enet_results |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  arrange(mean) |>
  head(10)
```

```{r}
#| label: enet-final

best_enet <- enet_results |> select_best(metric = "rmse")

final_enet_spec <- linear_reg(penalty = best_enet$penalty,
                              mixture = best_enet$mixture) |>
  set_engine("glmnet")

enet_wf <- workflow() |>
  add_model(final_enet_spec) |>
  add_recipe(reg_recipe)

enet_fit <- last_fit(enet_wf, split = music_split)
enet_fit |> collect_metrics()
```

---

## Model Comparison

| Method | Optimal λ | Mixture (α) | CV RMSE | Test RMSE |
|--------|-----------|-------------|---------|-----------|
| OLS (baseline) | — | — | — | 17.03 |
| Ridge | 10 | 0 (L2) | ~17.06 | 16.54 |
| Lasso | (tuned) | 1 (L1) | (see output) | (see output) |
| Elastic Net | (tuned) | (tuned) | (see output) | (see output) |

---

## Conclusions

All three regularization methods address overfitting in high-dimensional settings by penalizing coefficient magnitudes. For this music dataset:

- **Ridge regression** (λ = 10) improved on OLS, reducing test RMSE from 17.03 to 16.54. This demonstrates that even mild shrinkage helps when there are 72 predictors and 375 training observations.
- **Lasso** performs automatic variable selection, producing a sparse model where many audio features are zeroed out — useful for interpretability.
- **Elastic net** provides flexibility to balance both behaviors via the mixture parameter.

The key practical distinction: if **interpretability and feature selection** matter, lasso is preferred. If **predictive stability** with correlated predictors is the priority, ridge or elastic net are better choices.
