---
title: "Non-Linear Regression: Wage Prediction with Splines & Polynomials"
subtitle: "Fitting Natural Splines and Polynomial Models for Age Effects on Wage"
author: "Saroj Bhatta"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    highlight-style: github
    code-fold: false
    code-tools: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Many real-world relationships between predictors and outcomes are non-linear. Standard linear regression cannot capture these patterns without explicit feature engineering. This report explores two flexible non-linear methods — **natural splines** and **orthogonal polynomial regression** — for modeling the relationship between `age` and `wage`, with degrees of freedom tuned via 10-fold cross-validation. Partial dependence plots are used to visualize the estimated age effect.

---

## Setup

```{r}
#| label: load-packages

library(tidyverse)
library(tidymodels)
library(DALEXtra)
library(visdat)
```

---

## Dataset Overview & Missing Data

The wage dataset (adapted from ISLR) contains 3,000 observations and 11 variables describing worker characteristics and earnings in the mid-Atlantic region. The prediction target is `wage` (weekly earnings in thousands of dollars).

```{r}
#| label: load-data

load("data/wage.rda")
```

```{r}
#| label: fig-visdat
#| fig-cap: "Variable types and missing data in the wage dataset."

vis_dat(wage)
```

```{r}
#| label: fig-vismiss
#| fig-cap: "Missing data breakdown: maritl (9%) and education (3%) have missing values."

vis_miss(wage) +
  theme(axis.text.x = element_text(hjust = 0.5))
```

```{r}
#| label: missing-counts

colSums(is.na(wage))
```

The dataset has **11 variables** and **3,000 observations**. About 1.1% of data is missing:

- `maritl`: 264 missing (~9%)
- `education`: 88 missing (~3%)
- All other variables: complete

---

## Outcome Distribution

The analysis predicts `wage` from `age`, `health_ins`, `jobclass`, `education`, and `race`.

```{r}
#| label: fig-wage-dist
#| fig-cap: "Distribution of weekly wage. The right-skewed shape is consistent with wage inequality — most workers earn moderate wages with a long upper tail of high earners."

ggplot(data = wage, aes(x = wage)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 color = 'black', fill = 'skyblue') +
  geom_density(alpha = 0.5, fill = 'blue') +
  labs(
    title = "Distribution of Wage",
    x = "Wage ($000s per week)",
    y = "Density"
  )
```

---

## Preprocessing Recipe

Missing values are handled with `step_naomit(all_predictors())`, which removes rows with missing predictor values. Since only ~3% of observations are affected (driven by `education`), complete-case analysis introduces minimal bias and avoids the complexity of imputation.

```{r}
#| label: base-recipe

wage_recipe <- recipe(wage ~ age + health_ins + jobclass + education + race,
                      data = wage) |>
  step_naomit(all_predictors())
```

---

## Natural Splines

### Theory

Natural splines are piecewise polynomials joined at **knots**, with two additional constraints:

1. Continuity and smoothness at knot locations.
2. **Linear behavior beyond the boundary knots** — this prevents unstable oscillations at the extremes of the data (unlike unconstrained polynomials).

The **degrees of freedom** (df) parameter controls flexibility:

- df = k → k−1 internal knots placed at evenly spaced quantiles of the predictor
- Example: df = 3 → 2 internal knots at the 33.3rd and 66.7th percentiles of `age`

```{r}
#| label: spline-knot-example

# For df = 3: internal knots at the 33.3% and 66.7% quantiles of age
quantile(wage$age, probs = c(1/3, 2/3))
```

With df = 3, the knots fall at **age = 37 and age = 48**.

### Tuning Degrees of Freedom via 10-Fold CV

Rather than pre-specifying df, `tune()` is used inside the recipe to allow cross-validation to select the optimal value.

```{r}
#| label: spline-recipe

wage_recipe <- recipe(wage ~ age + health_ins + jobclass + education + race,
                      data = wage) |>
  step_naomit(all_predictors()) |>
  step_ns(age, deg_free = tune())
```

```{r}
#| label: spline-workflow

wf <- workflow() |>
  add_recipe(wage_recipe) |>
  add_model(linear_reg(engine = 'lm'))
```

```{r}
#| label: spline-tune

grid <- data.frame(deg_free = 1:10)

set.seed(142)
wage_folds <- vfold_cv(wage, v = 10)

tuned_results <- tune_grid(wf, resamples = wage_folds, grid = grid)

show_best(tuned_results, metric = "rmse")
```

The optimal degrees of freedom is **df = 6** with CV RMSE = **34.46**.

### Finalizing and Fitting the Spline Model

```{r}
#| label: spline-final-fit

best_param <- select_best(tuned_results, metric = 'rmse')

final_wf <- wf |> finalize_workflow(best_param)

final_fit <- fit(final_wf, data = wage)
```

### Partial Dependence Plot

A Partial Dependence Plot (PDP) shows the **marginal effect of age on predicted wage**, averaging over all other predictor values.

```{r}
#| label: spline-pdp-setup

source("ggplot_pdp.R")

model_explainer <- explain_tidymodels(
  final_fit,
  data  = wage,
  y     = wage$wage,
  label = 'Linear Regression with Age Spline'
)

age_profile_pdp <- model_profile(model_explainer, variables = 'age')
```

```{r}
#| label: fig-pdp-spline
#| fig-cap: "Partial dependence of predicted wage on age using the natural spline model (df = 6). Wages rise steeply in early career, plateau in middle age, then decline for older workers."

ggplot_pdp(age_profile_pdp, age) +
  labs(
    x = 'Age',
    y = 'Average Predicted Wage',
    title = 'Partial Dependence Profile: Natural Spline (df = 6)'
  )
```

The PDP reveals a **non-linear, inverted-U relationship** between age and predicted wage:

- Wages rise steeply from early career (~age 18) through middle age (~age 40)
- A plateau from approximately age 40–60
- A gradual decline for older workers (age > 60)

This pattern would be completely missed by a linear model.

---

## Polynomial Regression

As an alternative, orthogonal polynomial basis expansions of `age` are fit with degree tuned over 1–10 via 10-fold CV.

### Why Natural Splines vs. Polynomials?

- Polynomials force a single global function to fit all parts of the data range, which can cause **Runge's phenomenon** — extreme oscillations at the boundaries with high degrees.
- Natural splines use local piecewise functions with boundary constraints, producing smoother and more stable fits at extreme age values.

```{r}
#| label: poly-recipe

wage_recipe_poly <- recipe(wage ~ age + health_ins + jobclass + education + race,
                           data = wage) |>
  step_naomit(all_predictors()) |>
  step_poly(age, degree = tune())

wf_poly <- wf |> update_recipe(wage_recipe_poly)
```

```{r}
#| label: poly-tune

grid_poly <- data.frame(degree = 1:10)

tuned_poly_results <- tune_grid(wf_poly,
                                grid = grid_poly,
                                resamples = wage_folds)

show_best(tuned_poly_results, metric = 'rmse')
```

The best polynomial model uses **degree = 7** with CV RMSE = **34.47**.

```{r}
#| label: poly-final-fit

best_param_poly  <- select_best(tuned_poly_results, metric = 'rmse')

final_wf_poly  <- wf_poly |> finalize_workflow(best_param_poly)
final_fit_poly <- fit(final_wf_poly, data = wage)
```

```{r}
#| label: poly-pdp-setup

model_explainer_poly <- explain_tidymodels(
  final_fit_poly,
  data  = wage,
  y     = wage$wage,
  label = 'Linear Regression with Age Polynomial'
)

age_profile_poly <- model_profile(model_explainer_poly, variables = 'age')
```

```{r}
#| label: fig-pdp-poly
#| fig-cap: "Partial dependence of predicted wage on age using the 7th-degree polynomial model. The overall shape mirrors the spline, but with sharper oscillations at the boundaries."

ggplot_pdp(age_profile_poly, age) +
  labs(
    x = 'Age',
    y = 'Average Predicted Wage',
    title = 'Partial Dependence of Wage on Age (7th-Degree Polynomial)'
  )
```

Both models capture a similar non-linear age-wage relationship. However, the polynomial curve shows **sharper oscillations near age 70–80**, a known characteristic of high-degree polynomials beyond the range of dense training data.

---

## Model Comparison: Spline vs. Polynomial

| Method | Best Parameter | CV RMSE |
|--------|---------------|---------|
| Natural Spline | df = 6 | 34.46 |
| Polynomial | degree = 7 | 34.47 |

The natural spline is preferred:

1. **Nearly identical CV RMSE** — no predictive advantage to the polynomial.
2. **Boundary stability** — natural splines are constrained to be linear beyond the boundary knots, preventing the boundary oscillations seen in the polynomial PDP.
3. **Interpretability** — fewer effective parameters for the same fit quality.

---

## Does Regularization Help? Elastic Net + Spline

The optimal natural spline features were combined with elastic net regularization to test whether penalizing coefficients could further improve the model. Predictors are dummy-encoded and normalized for comparability.

```{r}
#| label: enet-recipe

wage_recipe_reg <- recipe(wage ~ age + health_ins + jobclass + education + race,
                          data = wage) |>
  step_naomit(all_predictors()) |>
  step_ns(age, deg_free = 6) |>           # best df from spline tuning
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

reg_model <- linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet")

wf_reg <- workflow() |>
  add_recipe(wage_recipe_reg) |>
  add_model(reg_model)
```

```{r}
#| label: enet-tune

grid_reg <- expand_grid(
  penalty = seq(0, 3, by = 0.5),
  mixture = seq(0, 1, by = 0.1)
)

tuned_reg_results <- tune_grid(wf_reg, resamples = wage_folds, grid = grid_reg)

show_best(tuned_reg_results, metric = "rmse")
```

```{r}
#| label: fig-enet-rmse
#| fig-cap: "Cross-validated RMSE versus penalty strength for elastic net across mixture values. RMSE rises as regularization increases — the unpenalized spline model is already optimal."

tuned_reg_results |>
  collect_metrics() |>
  filter(.metric == 'rmse') |>
  ggplot(aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_point() +
  geom_line() +
  labs(
    title = "Elastic Net: Cross-Validated RMSE vs Penalty (λ)",
    x     = "Penalty (λ)",
    y     = "Cross-Validated Mean RMSE",
    color = "Mixture (α)"
  )
```

The best regularized model achieves its minimum RMSE at **λ = 0** (no regularization), confirming that penalizing the spline coefficients does not improve prediction. As λ increases, RMSE consistently rises across all α values — the unpenalized natural spline model is already well-specified.

---

## Conclusions

This analysis demonstrates the practical advantages of non-linear modeling when a predictor has a curved relationship with the outcome:

1. **Non-linearity matters:** The age-wage relationship has a clear inverted-U shape that a linear model would miss entirely. Both splines and polynomials capture this effectively.
2. **Natural splines are preferred:** The spline model (df = 6, RMSE = 34.46) is essentially equivalent to the best polynomial model (degree = 7, RMSE = 34.47), but with better boundary behavior and fewer effective parameters.
3. **Regularization adds no value here:** The elastic net analysis confirms the spline model is not overfit — penalizing coefficients only increases prediction error.

The partial dependence plots provide an intuitive, model-based summary of the age effect that is consistent with economic intuition: wages rise through early career, plateau in middle age, and decline in later years.
