---
title: "Ensemble Methods for Sales Prediction"
subtitle: "Bagging, Random Forests, Gradient Boosting, Elastic Net & Model Stacking on the Carseats Dataset"
author: "Saroj Bhatta"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    highlight-style: github
    code-fold: false
    code-tools: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Ensemble methods combine multiple models to produce predictions that are typically more accurate and robust than any individual model. This report compares four approaches — **bagged decision trees**, **random forests**, **gradient boosted trees**, and **elastic net** — for predicting car seat `Sales` from 10 retail and product attributes. A **stacked ensemble** (model stacking) is then constructed to combine all four into a single optimized predictor.

---

## Setup

```{r}
#| label: load-packages

library(tidyverse)
library(tidymodels)
library(ISLR)
library(visdat)
library(stacks)
```

---

## Dataset Overview

The `Carseats` dataset from the ISLR package contains **400 observations** and **11 variables** simulating car seat sales at retail locations. The prediction target is `Sales` (unit sales in thousands).

```{r}
#| label: fig-visdat
#| fig-cap: "Variable types in the Carseats dataset: 8 numeric and 3 factor variables."

vis_dat(Carseats) +
  theme(axis.text.x = element_text(hjust = 0.5))
```

```{r}
#| label: fig-vismiss
#| fig-cap: "Missing data check — 0% missing values across all variables."

vis_miss(Carseats) +
  theme(axis.text.x = element_text(hjust = 0.5))
```

```{r}
#| label: data-structure

str(Carseats)
dim(Carseats)
```

There are **11 variables**: `Sales`, `CompPrice`, `Income`, `Advertising`, `Population`, `Price`, `ShelveLoc`, `Age`, `Education`, `Urban`, and `US`. The factor variables `ShelveLoc`, `Urban`, and `US` will be dummy-encoded during preprocessing. There are **no missing values**.

---

## Outcome Distribution

```{r}
#| label: fig-sales-dist
#| fig-cap: "Distribution of car seat Sales (thousands of units). The distribution is approximately bell-shaped, centered around 7–8 thousand units."

ggplot(data = Carseats, aes(x = Sales)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 color = 'black', fill = 'skyblue') +
  geom_density(alpha = 0.5, fill = 'blue') +
  labs(
    title = "Distribution of Sales",
    x = "Sales (thousands of units)",
    y = "Density"
  )
```

---

## Preprocessing Recipe & Shared Workflow

A single recipe is applied to all models. Dummy encoding handles the three factor variables; normalization ensures comparability across tree-based methods and elastic net (which is sensitive to predictor scale).

```{r}
#| label: recipe-workflow

sales_rec <- recipe(Sales ~ ., data = Carseats) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_predictors())

base_wf <- workflow() |> add_recipe(sales_rec)
```

### Cross-Validation Setup

5-fold CV is used to tune all models. The `control_resamples()` object with `save_pred = TRUE` and `save_workflow = TRUE` enables model stacking later — the stacks package requires access to held-out predictions from each candidate model.

```{r}
#| label: cv-setup

set.seed(7)
carseats_folds <- vfold_cv(Carseats, v = 5)

# Required for stacking: saves predictions and workflows from each fold
ctrl <- tune::control_resamples(
  save_pred     = TRUE,
  save_workflow = TRUE
)

# Tree count grid shared across all tree-based models
trees_grid <- expand_grid(trees = c(10, 25, 50, 100, 200, 300))
```

---

## Model 1: Bagged Decision Trees

**Bagging** (Bootstrap Aggregating) reduces variance by averaging predictions from multiple decision trees, each fit on a bootstrapped sample. Setting `mtry = ncol(Carseats) - 1 = 10` uses all predictors at every split — the defining characteristic of bagging versus random forests.

```{r}
#| label: bagging-tune

bag_spec <- rand_forest(
  trees = tune(),
  mtry  = ncol(Carseats) - 1,
  engine = 'ranger',
  mode   = 'regression'
)

bagged_wf <- base_wf |> add_model(bag_spec)

bagged_results <- tune_grid(bagged_wf,
                            grid      = trees_grid,
                            resamples = carseats_folds,
                            control   = ctrl)

show_best(bagged_results, metric = 'rmse')
```

```{r}
#| label: fig-bagging-rmse
#| fig-cap: "Mean CV RMSE versus number of trees for bagged decision trees. RMSE drops sharply from 10 to 50 trees, then stabilizes — adding more than 100 trees provides minimal further benefit."

ggplot(
  bagged_results |> collect_metrics() |> filter(.metric == 'rmse'),
  aes(x = trees, y = mean)
) +
  geom_point() +
  geom_line() +
  labs(
    y     = 'Mean RMSE',
    x     = 'Number of Trees',
    title = "RMSE Across Different Numbers of Trees: Bagged Model"
  )
```

**Best configuration:** 100 trees, CV RMSE = **1.54**. Performance plateaus after ~100 trees, with slight fluctuations between 100–300.

---

## Model 2: Random Forest

**Random forest** extends bagging by randomly sampling a subset of predictors at each split decision (`mtry < total predictors`). This decorrelates the individual trees, further reducing variance compared to bagging.

```{r}
#| label: rf-setup

set.seed(7)
carseats_folds <- vfold_cv(Carseats, v = 5)
```

```{r}
#| label: rf-tune

rf_spec <- rand_forest(trees = tune()) |>
  set_engine("ranger") |>
  set_mode("regression")

rf_wf <- base_wf |> add_model(rf_spec)

rf_tuned_results <- tune_grid(rf_wf,
                              grid      = trees_grid,
                              resamples = carseats_folds,
                              control   = ctrl)

show_best(rf_tuned_results, metric = "rmse")
```

```{r}
#| label: fig-rf-rmse
#| fig-cap: "Mean CV RMSE versus number of trees for random forest. Performance improves steadily through 300 trees, with the best RMSE at 100 trees."

ggplot(
  rf_tuned_results |> collect_metrics() |> filter(.metric == 'rmse'),
  aes(x = trees, y = mean)
) +
  geom_point() +
  geom_line() +
  labs(
    y     = 'Mean RMSE',
    x     = 'Number of Trees',
    title = "RMSE Across Different Numbers of Trees: Random Forest"
  )
```

**Best configuration:** 100 trees, CV RMSE = **1.69**.

> **Note on seed behavior:** `set.seed()` controls future randomness — once set and `carseats_folds` is created, that part is reproducible. However, `tune_grid()` internally consumes many random numbers, so the RNG state after bagging tuning differs from a fresh start. Re-setting the seed and recreating `carseats_folds` before each model ensures consistency.

---

## Model 3: Gradient Boosted Trees

**Gradient boosting** builds trees sequentially — each tree is fit to the residuals of the current ensemble (XGBoost implementation). Unlike bagging/RF which average independent trees, boosting learns iteratively. With `tree_depth = 1` (decision stumps) and `learn_rate = 0.1`, the model learns slowly and steadily.

```{r}
#| label: boost-setup

set.seed(7)
carseats_folds <- vfold_cv(Carseats, v = 5)
```

```{r}
#| label: boost-tune

boost_spec <- boost_tree(
  tree_depth = 1,
  learn_rate = 0.1,
  trees      = tune()
) |>
  set_engine("xgboost") |>
  set_mode("regression")

boost_wf <- base_wf |> add_model(boost_spec)

boost_tuned_results <- tune_grid(boost_wf,
                                 grid      = trees_grid,
                                 resamples = carseats_folds,
                                 control   = ctrl)

show_best(boost_tuned_results, metric = 'rmse')
```

```{r}
#| label: fig-boost-rmse
#| fig-cap: "Mean CV RMSE versus number of trees for gradient boosting. Unlike bagging/RF, RMSE continues declining monotonically through 300 trees — suggesting even more trees may further improve performance."

ggplot(
  boost_tuned_results |> collect_metrics() |> filter(.metric == 'rmse'),
  aes(x = trees, y = mean)
) +
  geom_point() +
  geom_line() +
  labs(
    y     = 'Mean RMSE',
    x     = 'Number of Trees',
    title = "RMSE Across Different Numbers of Trees: Boosted Tree Model"
  )
```

**Best configuration:** 300 trees, CV RMSE = **1.28**. Unlike bagging and RF which plateau, boosting continues improving — suggesting even more trees could reduce RMSE further.

---

## Model 4: Elastic Net (Penalized Regression)

Elastic net combines L1 and L2 regularization. Despite being a linear method, it performs surprisingly well on the Carseats data — evidence of strong linear structure in the predictors.

```{r}
#| label: enet-setup

set.seed(7)
carseats_folds <- vfold_cv(Carseats, v = 5)
```

```{r}
#| label: enet-tune

grid_enet <- expand_grid(
  penalty = seq(0, 0.1, by = 0.01),
  mixture = c(0, 0.5, 1)
)

penalized_spec <- linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet", path_values = seq(0, 0.1, by = 0.01))

elnet_wf <- base_wf |> add_model(penalized_spec)

elas <- tune_grid(elnet_wf,
                  grid      = grid_enet,
                  resamples = carseats_folds,
                  control   = ctrl)

show_best(elas, metric = 'rmse')
```

```{r}
#| label: fig-enet-rmse
#| fig-cap: "Penalized regression RMSE versus penalty strength across mixture values. Lower penalties and lasso (α=1) achieve best performance."

collect_metrics(elas) |>
  filter(.metric == 'rmse') |>
  ggplot(aes(x = penalty, y = mean,
             group = mixture, color = mixture)) +
  geom_point() +
  geom_line() +
  labs(
    x     = "Penalty (λ)",
    y     = "Mean RMSE",
    color = "Mixture (α)",
    title = "Penalized Regression: RMSE vs Penalty"
  )
```

**Best configuration:** penalty = 0.02, mixture = 1.0 (lasso), CV RMSE = **1.04**.

> Elastic net substantially outperforms all tree-based methods — the Carseats data has strong linear structure that penalized regression exploits efficiently. Boosting is powerful on complex non-linear problems, but elastic net often excels on medium-sized, correlated datasets with linear relationships.

---

## Individual Model Comparison

| Method | Best Configuration | CV RMSE |
|--------|-------------------|---------|
| Random Forest | 100 trees | 1.69 |
| Bagged Trees | 100 trees | 1.54 |
| Gradient Boosting | 300 trees | 1.28 |
| **Elastic Net** | λ=0.02, α=1 | **1.04** |

**Preferred individual model:** Elastic Net (RMSE = 1.04), followed by gradient boosting (1.28).

---

## Model Stacking: The Final Ensemble

A stacked ensemble is constructed using the `stacks` package. Stacking uses a meta-learner (here, regularized linear regression) to find the optimal **weighted combination** of predictions from all candidate models. The `blend_predictions()` step applies lasso regularization to select and weight the best candidates; `fit_members()` retrains the selected members on the full training data.

```{r}
#| label: stacking

ens <- stacks() |>
  add_candidates(bagged_results) |>
  add_candidates(rf_tuned_results) |>
  add_candidates(boost_tuned_results) |>
  add_candidates(elas) |>
  blend_predictions() |>
  fit_members()

ens
```

### Stack Members and Weights

The ensemble retained **5 members from 51 candidates** — all from the elastic net family. This confirms that the linear structure captured by penalized regression dominated the signal in the data; tree-based predictions added no incremental value beyond what elastic net already captured.

```{r}
#| label: stack-predictions

ens_preds <- ens |>
  predict(new_data = Carseats, members = TRUE) |>
  bind_cols(Carseats |> select(Sales))

# RMSE for ensemble and each retained member
ens_preds |>
  pivot_longer(cols = -Sales, names_to = "member", values_to = "prediction") |>
  group_by(member) |>
  rmse(truth = Sales, estimate = prediction)
```

```{r}
#| label: ensemble-rmse

ens_preds |> rmse(truth = Sales, estimate = .pred)
```

### Final Performance Summary

| Model | Best Config | CV RMSE |
|-------|------------|---------|
| Random Forest | 100 trees | 1.69 |
| Bagged Trees | 100 trees | 1.54 |
| Gradient Boosting | 300 trees | 1.28 |
| Elastic Net | λ=0.02, α=1 | 1.04 |
| **Stacked Ensemble** | 5 elastic net members | **1.01** |

The stacked ensemble achieves **RMSE = 1.01** — the best performance overall. The improvement over the best individual model (elastic net, 1.04) is modest but consistent with ensemble theory: combining complementary predictions from multiple models reduces error even when the best individual model already dominates.

---

## Conclusions

This analysis compared four ensemble and regularized regression methods for predicting car seat sales:

1. **Gradient boosting** was the best single tree-based method (RMSE = 1.28), consistently outperforming bagging (1.54) and random forest (1.69).
2. **Elastic net** unexpectedly dominated all tree methods (RMSE = 1.04). This is because the Carseats data has strong linear structure — `ShelveLoc`, `Price`, and `CompPrice` have largely linear relationships with `Sales`.
3. **Model stacking** provided a final marginal improvement (RMSE = 1.01) by optimally weighting elastic net variants, confirming that stacking is most beneficial when ensemble members are diverse and complementary.

The sole retention of elastic net models in the stack highlights an important practical lesson: **simpler, well-regularized linear models can outperform complex non-linear methods** when the underlying data structure is predominantly linear.
